Serum Biomarkers of Nutrition Pre and Post CFTR Modulator Use in Children with Cystic Fibrosis

### 2022-08-08 Edith Zemanick/CF Nutrition/cf_nutrition.Rmd
Update cf_nutrition.Rmd

```{r}
# Vitamins
vit = read_excel("./Edith Zemanick/CF Nutrition/Data_Raw/Report Vitamins_throughDecember2021_revised20220712.xlsx")
# Determine which modulator they were on
mods = read_excel("./Edith Zemanick/CF Nutrition/Data_Raw/Report CFTR Modulators_throughDecember2021.xlsx")
mods = mods[mods$Modulator %in% c('Kalydeco','Trikafta'),]
mods = mods %>% select(`Patient ID`,Modulator,`Earliest Date in Registry`) %>% 
  pivot_wider(id_cols = `Patient ID`,names_from = Modulator,
              values_from = `Earliest Date in Registry`)
colnames(mods)[2:ncol(mods)] = paste0(colnames(mods)[2:ncol(mods)]," Date")
mods$`First Mod Date` = pmin(mods$`Trikafta Date`,mods$`Kalydeco Date`,na.rm = T)
mods$`Second Mod Date` = pmax(mods$`Trikafta Date`,mods$`Kalydeco Date`)
# Add pancreatic status
pancreatic = read_excel("./CF Info/Sweat Tests from Internal Database_genotype class_withPancStatus.xlsx")
pancreatic = pancreatic %>% rename(`Patient ID` = PatientID) %>% select(`Patient ID`,PancreaticStatus)
# FEV1 and BMI
fev = read_excel("./Edith Zemanick/CF Nutrition/Data_Raw/Report_PFTs_ThroughDecember2021.xlsx")
fev = fev %>% select(PatientID,Date,BMI,`FEV1 % pred`) %>% rename(`Patient ID` = PatientID)
# Merge
df = left_join()
# Merge
df = full_join(vit,fev,by = c("Patient ID", "Date")) %>% filter(`Patient ID` %in% mods$`Patient ID`)
df = left_join(df,mods,by = "Patient ID")
df = left_join(df,pancreatic,by = "Patient ID")
# Time relative to first mod
df$Time = as.numeric(difftime(df$Date,df$`First Mod Date`,units = "days"))/365.25
# Calculate spline things
df$c1 = (df$Time-0)*(df$Time>=0)
df$c2 = as.numeric(difftime(df$Date,df$`Second Mod Date`,units = "days"))/365.25
df$c2 = df$c2*(df$Date >= df$`Second Mod Date`)
# Convert to numeric
df$VitaminA_Retinol = suppressWarnings(as.numeric(df$VitaminA_Retinol))
# Select columns
df = df %>% filter(!is.na(VitaminA_Retinol)) %>%
  select(`Patient ID`,`First Mod Date`,`Second Mod Date`,
         Date,Time,c1,c2,VitaminA_Retinol) %>%
  rename(`Vitamin A` = VitaminA_Retinol)
# Write for presentation
write.csv(df,file = "~/GitHub/CF/StatNet/data.csv",row.names = F,na = "")
```

```{r}
# Counts from Elin
# patients with CF in our clinical database from 2008- December 2021: 740 total, 474 in our dataset
ncdb = 740
# of patients treated with ivacaftor from 2008-2021: 58 total, 47 in dataset
# of patients treated with trikafta (ETI) from 2008-2021: 228 total, 150 in dataset 
# of patients treated with ivacaftor followed by ETI – these patients may be counted in both groups: 22 total, 17 in dataset
# Tim counts
df$Group = apply(df,1,function(r){
  if(!is.na(r["Kalydeco Date"]) & is.na(r["Trikafta Date"])){
    g = "Kalydeco"
  } else if (is.na(r["Kalydeco Date"]) & !is.na(r["Trikafta Date"])){
    g = "Trikafta"
  } else if (!is.na(r["Kalydeco Date"]) & !is.na(r["Trikafta Date"])){
    g = "Both"
  }
  return(g)
})
# "Our dataset"
sweat = read_excel("./Edith Zemanick/CF Nutrition/Data_Raw/Report Electrolytes_throughDecember2021.xlsx")
df = df %>% filter(`Patient ID` %in% sweat$`Patient ID`)
groups = df %>% group_by(`Patient ID`) %>% filter(row_number()==1)
# Two nutrition measures before first mod start
two_measures = df %>% 
  mutate(before_first = Date < `First Mod Date` & !is.na(VitaminA_Retinol)) %>% 
  group_by(`Patient ID`,Group) %>% count(before_first) %>% 
  filter(before_first == T,n >= 1)
# At least one nutritional measure and one weight/height measure >= 3 months after first
after_measure = df %>% 
  mutate(after_first = (Date >= `First Mod Date` %m+% months(3)) & 
           !is.na(VitaminA_Retinol)) %>%
  group_by(`Patient ID`,Group) %>% count(after_first) %>%
  filter(after_first == T,n >= 1)
# Write merged df - filter to just Vitamin A values
t = df %>% 
  mutate(`Pre/Post First Mod` = ifelse(Date > `First Mod Date`,"Post","Pre")) %>%
  filter(!is.na(VitaminA_Retinol)) 
write.csv(t,file = "./Edith Zemanick/CF Nutrition/Data_Cleaned/merged_data.csv",
          row.names = F,na="")
write.csv(mods,file = "./Edith Zemanick/CF Nutrition/Data_Cleaned/modulators.csv",
          row.names = F,na="")
```

@@

<!-- # Participant Characteristics -->

<!-- For each person, we calculated the maximum amount of time from the earliest nutrition measure before modulator start to start date, and the amount of time from modulator start date to most recent nutrition measure. The ranges of these values are presented below ("Years to start" and "Years From start," respectively). -->

<!-- ```{r results='asis'} -->
<!-- df$Sex[df$Sex == ""] = NA -->
<!-- demo = df %>% group_by(Patient.ID) %>% fill(Sex,.direction = "downup") %>% -->
<!--   filter(row_number() == 1) -->
<!-- t1 = tableby(first_mod ~ Age.at.Start..years. + Sex + genoRisk + -->
<!--                vit_measures_after + vit_measures_before + -->
<!--                notest(max_years_to_start,"min","max") + -->
<!--                notest(max_years_from_start,"min","max"), -->
<!--              data = demo, -->
<!--              control = tableby.control(numeric.stats=c("Nmiss", "median", "q1q3"),)) -->
<!-- summary(t1,test = F, -->
<!--         labelTranslations = -->
<!--           list(Age.at.Start..years. = "Age at Start (years)", -->
<!--                genoRisk = "Genotype risk group", -->
<!--                vit_measures_after = "No. of vitamin measures post", -->
<!--                vit_measures_before = "No. of vitamin measures pre", -->
<!--                max_years_to_start = "Years to start",max_years_from_start = "Years From start")) -->
<!-- ``` -->

>>>>

```{r}
# Vitamins
vit = read_excel("./Edith Zemanick/CF Nutrition/Data_Raw/Report Vitamins_throughDecember2021_revised20220712.xlsx")
# Determine which modulator they were on
mods = read_excel("./Edith Zemanick/CF Nutrition/Data_Raw/Report CFTR Modulators_throughDecember2021.xlsx")
mods = mods %>% filter(Modulator %in% c('Kalydeco','Trikafta')) %>%
  select(`Patient ID`,Modulator,`Earliest Date in Registry`) %>% 
  pivot_wider(id_cols = `Patient ID`,names_from = Modulator,
              values_from = `Earliest Date in Registry`)
colnames(mods)[2:ncol(mods)] = paste0(colnames(mods)[2:ncol(mods)]," Date")
mods$`First Mod Date` = pmin(mods$`Trikafta Date`,mods$`Kalydeco Date`,na.rm = T)
mods$`Second Mod Date` = pmax(mods$`Trikafta Date`,mods$`Kalydeco Date`)
# Add pancreatic status
pancreatic = read_excel("./CF Info/Sweat Tests from Internal Database_genotype class_withPancStatus.xlsx")
pancreatic = pancreatic %>% rename(`Patient ID` = PatientID) %>% select(`Patient ID`,PancreaticStatus)
# FEV1 and BMI
fev = read_excel("./Edith Zemanick/CF Nutrition/Data_Raw/Report_PFTs_ThroughDecember2021.xlsx")
fev = fev %>% select(PatientID,Date,BMI,`FEV1 % pred`) %>% rename(`Patient ID` = PatientID)
# Merge
df = full_join(vit,fev,by = c("Patient ID", "Date")) %>% filter(`Patient ID` %in% mods$`Patient ID`)
df = left_join(df,mods,by = "Patient ID")
df = left_join(df,pancreatic,by = "Patient ID")
```

```{r}
# Counts from Elin
# patients with CF in our clinical database from 2008- December 2021: 740 total, 474 in our dataset
ncdb = 740
# of patients treated with ivacaftor from 2008-2021: 58 total, 47 in dataset
# of patients treated with trikafta (ETI) from 2008-2021: 228 total, 150 in dataset 
# of patients treated with ivacaftor followed by ETI – these patients may be counted in both groups: 22 total, 17 in dataset
# Tim counts
df$Group = apply(df,1,function(r){
  if(!is.na(r["Kalydeco Date"]) & is.na(r["Trikafta Date"])){
    g = "Kalydeco"
  } else if (is.na(r["Kalydeco Date"]) & !is.na(r["Trikafta Date"])){
    g = "Trikafta"
  } else if (!is.na(r["Kalydeco Date"]) & !is.na(r["Trikafta Date"])){
    g = "Both"
  }
  return(g)
})
# "Our dataset"
sweat = read_excel("./Edith Zemanick/CF Nutrition/Data_Raw/Report Electrolytes_throughDecember2021.xlsx")
df = df %>% filter(`Patient ID` %in% sweat$`Patient ID`)
groups = df %>% group_by(`Patient ID`) %>% filter(row_number()==1)
# Filter any measures within 3 months of modulator start
df = df %>% 
  filter(Date < `First Mod Date` - days(90) | 
           ((Date > `First Mod Date` + days(90)) & (Date < `Second Mod Date` - days(90))) |
           (Date > `Second Mod Date` + days(90) | is.na(`Second Mod Date`)))
# Two nutrition measures before first mod start
two_measures = df %>% 
  mutate(before_first = Date < `First Mod Date` & 
           (!is.na(VitaminA_Retinol) | !is.na(`25OH-Vitamin D`) | !is.na(`Vitamin E Alpha`))) %>% 
  group_by(`Patient ID`,Group) %>% count(before_first) %>% 
  filter(before_first == T,n >= 1)
# At least one nutritional measure and one weight/height measure >= 3 months after first
after_measure = df %>% 
  mutate(after_first = (Date >= `First Mod Date` %m+% months(3)) & 
           (!is.na(VitaminA_Retinol) | !is.na(`25OH-Vitamin D`) | !is.na(`Vitamin E Alpha`))) %>%
  group_by(`Patient ID`,Group) %>% count(after_first) %>%
  filter(after_first == T,n >= 1)
# Filter
df = df %>% filter(`Patient ID` %in% two_measures$`Patient ID` & 
                     `Patient ID` %in% after_measure$`Patient ID`)
# Get most recent pre-/post-first modulator values
# Time from first mod
df$Time = as.numeric(difftime(df$Date,df$`First Mod Date`,units = "days"))
# Before or after start
df$Before_After = cut(df$Time,c(-Inf,0,Inf),labels = c("Before","After"),right = T)
t = df %>% filter(abs(Time) <= (3*365)) %>% # Within 3 years of start
  group_by(`Patient ID`,Before_After) %>% 
  arrange(abs(Time),.by_group = T) %>% # Sort so that the the first non-NA value is the closest to modulator start
  summarise(`Vitamin A` = first(na.omit(VitaminA_Retinol)),
            `Vitamin D` = first(na.omit(`25OH-Vitamin D`)),
            `Vitamin E Alpha` = first(na.omit(`Vitamin E Alpha`)),
            `Vitamin E Gamma` = first(na.omit(`Vitamin E Gamma`)),
            BMI = first(na.omit(BMI)),.groups = "drop")
```

@@

# Participant Characteristics

For each person, we calculated the maximum amount of time from the earliest nutrition measure before modulator start to start date, and the amount of time from modulator start date to most recent nutrition measure. The ranges of these values are presented below ("Years to start" and "Years From start," respectively).

```{r results='asis'}
# Import demo info and add

df$Sex[df$Sex == ""] = NA
demo = df %>% group_by(Patient.ID) %>% fill(Sex,.direction = "downup") %>%
  filter(row_number() == 1)
t1 = tableby(first_mod ~ Age.at.Start..years. + Sex + genoRisk +
               vit_measures_after + vit_measures_before +
               notest(max_years_to_start,"min","max") +
               notest(max_years_from_start,"min","max"),
             data = demo,
             control = tableby.control(numeric.stats=c("Nmiss", "median", "q1q3"),))
summary(t1,test = F,
        labelTranslations =
          list(Age.at.Start..years. = "Age at Start (years)",
               genoRisk = "Genotype risk group",
               vit_measures_after = "No. of vitamin measures post",
               vit_measures_before = "No. of vitamin measures pre",
               max_years_to_start = "Years to start",max_years_from_start = "Years From start"))
```
